% Postdoc development plan due 2022.07.01.

% # Fast Track Initiative
%
% Do you have a good idea that might lead to a new project?
%
% The Fast-track initiatives (FTI) support focused, short-term research
% activities with the aim of achieving concise result(s).
%
% The call goes out every autumn, and deadline for submission is usually
% December 1. Calls for FTI will be announced by email to BCCR members in due
% time before the deadline.
%
% The Fast Track Initiative is set up to support pilot studies, proposals,
% synergy works and other intitiatives. We have several large projects today
% that started out as a fast track initiative.
%
% Limits to application: Maximum 500 kNOK
%
% Who can apply: Anyone affiliated with the Bjerknes Centre for Climate
% Research, resident in Bergen
%
% ## Criteria
%
% - Must target at least one of Bjerknes strategic research areas
% - Promote the integration of the Bjerknes consortium
% - Must have clearly identifiable outcome
% - Must be short-lived and completed within 1 year of the grant´s commencement
% - Proposals should not exceed one A4 page including a description of the
%   activity, objective(s), budget, timeline and outcome(s). If salaries are
%   applied for, the person shall be identified and an estimate of the salary
%   costs must be given by an economy consultant at one of the partner
%   institutions and according to the agreement on cost per hour for the SKD
%   period 2021-2026.
%
% ## Some examples of eligible activities
%
% Note that this list is not exhaustive:
%
% - Blue-sky research: identify or conduct exploratory activities including
%   laboratory work, for new, cross-cutting research relevant to Bjerknes
% - Preparation of synthesis/integration research articles or large research
%   proposals
% - Up to 3 month´s salary for a recently graduated master student to complete
%   a manuscript for publication at salary rate 42
%
% Non-eligible activities or expenses:
%
% - Running costs such as travels, conference fees, publication costs, etc.
% - Salaries of non-Bjerknes personnel
% - Salaries for PhD students before thesis is submitted
% - Research visits to Bergen. These must be applied to the Bjerknes Visiting
%   Fellow programme

% vanilla article for latex2rtf
\newif\iflatextortf
\iflatextortf
  \documentclass{article}
  \renewcommand{\cvdoubleitem}[4]{#1: #2; #3: #4}
  \renewcommand{\href}[2]{\htmladdnormallink{\color{blue}\underline{#2}}{#1}}

% else moderncv just looks cool
\else
  \documentclass[11pt,a4paper,sans,colorlinks]{moderncv}
  \name{Julien}{Seguinot}  % only used in metadata
  \moderncvstyle[left]{casual}  % casual (default), classic, oldstyle or banking
  \moderncvcolor{blue} % blue (default), orange, green, red, purple, grey or black

\fi

% latex packages
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[bottom=0.08\paperheight,top=0.08\paperheight]{geometry}

% moderncv section numbers
% \newcounter{secnumber}
% \renewcommand\sectionstyle[1]{{
%   \refstepcounter{secnumber}\sectionfont\textcolor{color1}{\thesecnumber.~#1}}}

% custom commands
\newcommand{\todo}[1]{{\color{red}\emph{Todo}: #1}}

% document properties
\title{Development plan for Postdoctoral Research Fellow}

% ======================================================================
\begin{document}
\setlength{\parskip}{0.5\baselineskip}
% ======================================================================

    \begin{center}
      \textbf{Development plan for Postdoctoral Research Fellow}\\
      Julien~Seguinot \\
      Department of Biological Sciences and
      Bjerknes Centre for Climate Research
    \end{center}

    Most glaciers lead lives of constant stress, so they relax by slowly
    gliding on their beds.

    A critical glaciologic variable, stress is almost never measured directly.
    Empirical glacier mechanics instead rely on controlled laboratory
    experiments on ice and indirect measurements of the stress response in
    glacier surface velocities and englacial tilt rates. Nevertheless, there is
    no lack of observational evidence that crystal orientation, ice water
    content and impurities may complicate laboratory-derived laws in nature.
    Direct stress measurements in glacier may sound more like a theoretician's
    dream than reality, but if feasible, they could be a drastic accelerator in
    understanding glacier hazards and ice-ocean interactions.

    I have preliminary data that hints at their possibility. I hereby apply for
    funds to analyse and publish these data: an accidental, three-year record
    of englacial stress from water-pressure sensors frozen into a Greenlandic
    tidewater glacier.

% ----------------------------------------------------------------------
\section{Existing results}
% ----------------------------------------------------------------------

    Between summers 2014 and 2017, I was part of a Swiss-Japanese team of that
    deployed and maintained ice-borehole instruments on Bowdoin Glacier, a
    tidewater glacier in north-western Greenland. These instruments included
    multi-sensor digital units, each equipped with a piezometer to measure
    overhead water depth and thus its location in the hotwater-drilled
    borehole. While these water-pressure sensors were not intended for solid
    stress measurements, they continued to record pressure changes after the
    complete refreezing of the boreholes and the stabilisation of ice
    temperatures well below the pressure melting point. Unexpectedly, that
    record is not just noise.

    During Spring 2021 (then unemployed), and with guidance from my field
    colleague and glacier seismologist Evgeniy Podolskiy, I analysed the stress
    timeseries via fourier transform, rolling window spectrograms, wavelet
    transforms, cross-correlation and rolling cross-correlation.
    All sensors recorded in-phase pressure changes with 12-hour, 24-hour and
    14-day periodicities, corresponding to the local tidal signature.
    The amplitude of these pressure variations is only an order of magnitude
    lower than the tidal amplitude measured at sea. However, the pressure is
    anticorrelated with the tides, and shows a delay of one to two hours, so
    that maximum stress occurs a little after low tide.

% ----------------------------------------------------------------------
\section{Proposed work}
% ----------------------------------------------------------------------

    While we have enough figures for a paper, we still lack a solid
    interpretation for the phase relationship between oceanic tides and stress
    oscillations. We aim to seek an explanation in Bowdoin Glacier's surface
    velocity record. During the same observation period, Shin Sugiyama and
    colleagues collected GPS timeseries from several parts of the glacier. At
    least some of these records have a subtle tidal signature (overprinted by
    diurnal summer speed-ups).

    I hereby apply for funds to travel to Sapporo to analyze the spectral
    content of Bowdoin GPS signals, interpret the englacial stress timeseries,
    and compile the results in a publication.

    TODO: add ref to paper, blog.

% ----------------------------------------------------------------------
\section{Budget}
% ----------------------------------------------------------------------

    \begin{itemize}
        \item Three months of my own salary (135000 NOK?)
        \item Return travels to Sapporo (20000 NOK)
        \item Three months university accomodation (15000 NOK?)
        \item Funding for an open-access publication (20000 NOK)
    \end{itemize}

% ======================================================================
\end{document}
% ======================================================================
